---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---
## Loading libraries
 
```{r}
library (plyr)
library (lattice)
```

## Loading and preprocessing the data

Loading the Activity.csv data provided for the assignment.

```{r}
 data <- read.csv("activity.csv")
 str(data)
```

Data looks clean for now. Data will be formatted as required in future steps

## What is mean total number of steps taken per day?

Making a histogram of the total number of steps taken each day
```{r}
totalSteps <- tapply(data$steps, data$date, sum)
hist(totalSteps, breaks = 10)

```

Calculating the mean and median total number of steps taken per day
```{r}
mean(totalSteps, na.rm = TRUE)

```
```{r}
median(totalSteps, na.rm = TRUE)

```


## What is the average daily activity pattern?

Making a time series plot (i.e. type = "l") of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all days (y-axis)
```{r}
summarizedData <- ddply (data, .(interval), summarize, meanSteps = mean(steps, na.rm=TRUE))
 plot(summarizedData$interval, summarizedData$meanSteps, type ='l')

```

Calculating Which 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps.
```{r}
maxInterval <- summarizedData[summarizedData$meanSteps==max(summarizedData$meanSteps),]
maxInterval

```

Answer: 835


## Imputing missing values

Calculating the total number of missing values in the dataset (i.e. the total number of rows with NAs)
```{r}
sum(is.na(data$steps))
```
Strategy for filling in all of the missing values in the dataset:

The strategy is to use  the mean for that 5-minute interval and Creating  a new dataset that is equal to the original dataset but with the missing data filled in.
```{r}
newData <- data
newData$steps[is.na(newData$steps)] <- tapply(X=data$steps,INDEX=data$interval,FUN=mean,na.rm=TRUE)
head(newData)
```
Making a histogram of the total number of steps taken each day and Calculating the mean and median total number of steps taken per day, to illustrate the difference w.r.t original dataa
```{r}
 newSummary <- ddply (newData, .(interval), summarize, meanSteps = mean(steps, na.rm=TRUE))
plot(newSummary$interval, newSummary$meanSteps, type ='h')

summary(newSummary)

```


## Are there differences in activity patterns between weekdays and weekends?
Creating a new factor variable in the dataset with two levels -- "weekday" and "weekend" indicating whether a given date is a weekday or weekend day.

```{r}
days <- weekdays(as.Date(newData$date) )
newData$dayTypes <- factor (days, levels <- c ('weekday', 'weekend'))
newData$dayTypes[] <- 'weekday'
 newData$dayTypes[days %in% c ('Saturday', 'Sunday')] <- 'weekend'
 patternSummary  <- ddply (newData,.(interval, dayTypes),  summarize,                           stepMean = mean(steps, na.rm=TRUE))

```



Making a panel plot containing a time series plot (i.e. type = "l") of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all weekday days or weekend days (y-axis). The plot should look something like the following, which was created using simulated data:

```{r}
xyplot (stepMean ~ interval | dayTypes, 
      data=patternSummary,
         type='l',
       layout=c (1, 2))
```

